\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-02-15 doc/publications/plos16/plos16.tex
changes to conclustion

Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
large number of areas including variable selection in high dimensions. They show that there is a sharp disjunction
between the cases where significant variables may be recovered with a high accuracy by procedures like stepwise variable
selection, and the cases where they can not be recovered. The boundary is shown in
Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdeterminedness,
$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of significant variables). The theoretical
boundary is based on arguments from combinatorial geometry. Above the phase-transition line variable recovery is still
possible by a combinatorial approach such as all-subsets variable selection.

@@

\cite{Donoho.and.Stodden.2006} investigate the behavior of a number of
regression approaches for variable selection in a small simulation.  They consider a regression probem with 
$\underset{n\times p}{X}\sim N(0,1)$ with $p$ fixed at 200 and $n$ variable.
They  set $\beta(1:k) \sim U(1,100)$ and $\beta((k+1):p) =0$. Then  $y= X\beta + \epsilon$, $\epsilon \sim
N(0,\sqrt{16})$. They evaluate variable selection by the error measure 
$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}$$

They consider a number of variable selection methods, including a false discovery rate criteria. This involves adding
the variable with the maximum $t$-value to the linear model if the $p$-value is less than 
$25(\text{number of  terms  currently  in  the  model})/(\text{total  number  of  variables})$.  
See Fig~\ref{figure:error_Stodden_FDR.png} for the error measure and
figure \ref{figure:rbo_Stodden_FDR.png} for the RBO, comparing the ranking (i.e. values) of $\hat{\beta}$ and $\beta$.
It apparent that there is a marked drop in the accuracy of the variable recovery in line with the pediction of the
Donoho-Tanner phase transition.

\begin{figure}[tbhp] 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/k.png}
      \caption{$k$}
      \label{figure:k.png}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/n.png}
      \caption{$n$}
      \label{figure:n.png}
    \end{subfigure} 
    \caption{The simulation in \protect{\cite{Donoho.and.Stodden.2006}} considers $\delta = n/p$ and $\rho =k/n$. Here we plot the
      space of $\{ \delta, \rho$\}, colored  by the paramters $n$ and $k$}
\end{figure}

 \begin{figure}[tbhp] 
%   \begin{adjustwidth}{-1.00in}{0in}
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/error_Stodden_FDR.png}
       \caption{The error measure.}
       \label{figure:error_Stodden_FDR.png}
%       \vspace{4ex}
     \end{subfigure} 
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/rbo_Stodden_FDR.png}
       \caption{The RBO measure}
       \label{figure:rbo_Stodden_FDR.png}
     \end{subfigure} 
     \caption{Forward stepwise variable selction with a false discovery rate stopping criteria.}
       \label{figure:error_Stodden_FDR.png}
%   \end{adjustwidth}
 \end{figure}

 \begin{figure}[tbhp] 
%   \begin{adjustwidth}{-1.00in}{0in}
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/ranger_error_Stodden_simulation.png}
       \caption{Stepwise with FDR stopping criteria. $p=200$, $n$ and $k$ variable. }
       \label{figure:ranger_error_Stodden_simulation.png}
       \vspace{4ex}
     \end{subfigure} 
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/ranger_rbo_Stodden_simulation.png}
       \caption{RBO}
       \label{figure:ranger_rbo_Stodden_simulation.png}
     \end{subfigure} 
     \caption{}
     \label{figure:ranger_rbo_Stodden_simulation.png}
%   \end{adjustwidth}
 \end{figure}

We have seen no work on the question of such a phase transition in the case of classification problems. Still it amy be
instructive to consider the implications of the pahse transition for our work. As \cursedforest is designed for
extremely large numbers of variables it is likely to be operating in difficult regions of the figure where the ratio
$\delta = n/p$ is small. In the simulated example in section \ref{section:synthetic_data} the underdeterminedness $\delta = 0.002$ and the
sparsity $\rho =2\times 10^{-6}$

We have used a random forest on the same simulation as figure \ref{figure:error_Stodden_FDR.png}.  The Donoho-Tanner phase transition arises in
recovering the $\beta$ in data generated by a linear model. However, in a decision tree (random forest) we are fitting a
non-linear model and there is no notion of estimating the $\beta$. Becsaue of this we have evaluted the performance of
the random forest using the RBO measure on the variable importance.

We note that while a random forest is a long way from an all-seubsets search, it is a limited combinatorial search of
the feature space. As such it may perform outside of the bounds of the phase transition. 
  combinatorial search.

Fig~\ref{figure:ranger_rbo_Stodden_simulation.png} shosw the OOB prediction error and the RBO error for a random forest with 10000 trees and the \mtry value set
to the default for a regression problem (depending on $n$). It shows no evidence of following the shape of the
Donoho-Tanner phase transition.

@@

>>>>

Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
large number of areas including variable selection in high dimensions. They show that there is a sharp disjunction
between the cases where significant variables may be recovered with a high accuracy by procedures like stepwise variable
selection, and the cases where they can not be recovered. The boundary is shown in
Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdetermination,
$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of significant variables).  Above the
phase-transition line variable recovery is still possible by a combinatorial search approach such as all-subsets
variable selection.  The theoretical boundary is based on arguments from combinatorial geometry.

@@

\cite{Donoho.and.Stodden.2006} investigate the behavior of a number of
regression approaches for variable selection in a small simulation.  They consider a regression problem with 
$\underset{n\times p}{X}\sim N(0,1)$ with $p$ fixed at 200 and $n$ variable.
They  set $\beta(1:k) \sim U(1,100)$ and $\beta((k+1):p) =0$, then  $y= X\beta + \epsilon$, $\epsilon \sim
N(0,\sqrt{16})$. They evaluate variable selection by the error measure 
$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}$$

They consider a number of variable selection methods, including a false discovery rate criteria. This involves adding
the variable with the maximum $t$-value to the linear model if the $p$-value is less than 
$25(\text{number of  terms  currently  in  the  model})/(\text{total  number  of  variables})$.  
See Fig~\ref{figure:error_Stodden_FDR.png} for the error measure and
figure \ref{figure:rbo_Stodden_FDR.png} for the RBO, comparing the ranking (i.e. values) of $\hat{\beta}$ and $\beta$.
It apparent that there is a marked drop in the accuracy of the variable recovery in line with the prediction of the
Donoho-Tanner phase transition.

 \begin{figure}[tbhp] 
%   \begin{adjustwidth}{-1.00in}{0in}
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/error_Stodden_FDR.png}
       \caption{The error measure.}
       \label{figure:error_Stodden_FDR.png}
%       \vspace{4ex}
     \end{subfigure} 
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/rbo_Stodden_FDR.png}
       \caption{The RBO measure}
       \label{figure:rbo_Stodden_FDR.png}
     \end{subfigure} 
     \caption{Forward stepwise variable selection with a false discovery rate stopping criteria.}
       \label{figure:error_Stodden_FDR.png}
%   \end{adjustwidth}
 \end{figure}

 \begin{figure}[tbhp] 
%   \begin{adjustwidth}{-1.00in}{0in}
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/ranger_error_Stodden_simulation.png}
       \caption{The OOB error rate.}
       \label{figure:ranger_error_Stodden_simulation.png}
       \vspace{4ex}
     \end{subfigure} 
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/ranger_rbo_Stodden_simulation.png}
       \caption{The RBO measure}
       \label{figure:ranger_rbo_Stodden_simulation.png}
     \end{subfigure} 
     \caption{The simulated data of \protect{\cite{Donoho.and.Stodden.2006}} with a random forest model.}
     \label{figure:ranger_rbo_Stodden_simulation.png}
%   \end{adjustwidth}
 \end{figure}

We have seen no work on the question of such a phase transition in the case of classification problems. Still it may be
instructive to consider the implications of the phase transition for our work. As \cursedforest is designed for
extremely large numbers of variables it is likely to be operating in difficult regions of the figure where the ratio
$\delta = n/p$ is small. In the simulated example in section \ref{section:synthetic_data} the underdetermination $\delta = 0.002$ and the
sparsity $\rho =2\times 10^{-6}$

We have used a random forest on the same simulation as figure \ref{figure:error_Stodden_FDR.png}.  The Donoho-Tanner phase transition arises in
recovering the $\beta$ in data generated by a linear model. However, in a decision tree (random forest) we are fitting a
non-linear model and there is no notion of estimating the $\beta$. Because of this we have evaluated the performance of
the random forest using the RBO measure on the variable importance.

We note that while a random forest is a long way from an all-subsets search, it is a limited combinatorial search of
the feature space. As such it may perform outside of the bounds of the phase transition. 
  combinatorial search.

Fig~\ref{figure:ranger_rbo_Stodden_simulation.png} shows the OOB prediction error and the RBO error for a random forest with 10000 trees and the \mtry value set
to the default for a regression problem (depending on $n$). It shows no evidence of following the shape of the
Donoho-Tanner phase transition.

@@

\begin{figure}[tbhp] 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/k.png}
      \caption{$k$}
      \label{figure:k.png}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/n.png}
      \caption{$n$}
      \label{figure:n.png}
    \end{subfigure} 
    \caption{The simulation in \protect{\cite{Donoho.and.Stodden.2006}} considers $\delta = n/p$ and $\rho =k/n$. Here we plot the
      space of $\{ \delta, \rho$\}, colored  by the parameters $n$ and $k$}
\end{figure}
